---
title: "Non-contiguous hex-cartograms"
subtitle: "Local Authority and Middle-layer Super Output Area exmaples"
author: "Tom Rushby"
date: "25/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r requiredPackages}
library(curl)
library(sf)
library(tidyverse)
library(ggplot2)
```


## Introduction

This example follows an example (and re-uses code) by [@VictimOfMaths](https://github.com/VictimOfMaths/Maps/blob/master/WFHCartogram.R) using hex-map data from [House of Commons Library](https://github.com/houseofcommonslibrary/uk-hex-cartograms-noncontiguous/) by [@carlbaker](https://twitter.com/carlbaker).

The first example we will look at is plotting a non-contiguous hex cartogram of Local Authority areas.

## Local Authority areas

```{r downloadHexMaps}
ltla <- tempfile()
source <- ("https://github.com/houseofcommonslibrary/uk-hex-cartograms-noncontiguous/raw/main/geopackages/LocalAuthorities-lowertier.gpkg")
ltla <- curl_download(url=source, destfile=ltla, quiet=FALSE, mode="wb")
```

The file downloaded is a GeoPackage (.gpkg) file, an SQLite Database container  (see http://www.geopackage.org/guidance/getting-started.html for more info.). We can examine the layers contained using the `st_layers()` command from the sf package.

```{r examineGeoLayers}
st_layers(ltla)
```

Next we extract the layers we want using the `st_read` command. [Alternative methods](https://olalladiaz.net/blog/2018/11/02/working-with-gpkg-r/) are available.

```{r extractGeoLayers}
Background <- st_read(ltla, layer="7 Background")

Areas <- st_read(ltla, layer="4 LTLA-2019")

Groups <- st_read(ltla, layer="2 Groups")

Group_labels <- st_read(ltla, layer="1 Group labels") %>% 
  mutate(just=if_else(LabelPosit=="Left", 0, 1))
```

And finally we can make a plot ... note the `Groups` geometry provides County-level grouping of local authorities and an outline (appears bold in the plot below). `Areas` provides the local authority outlines.

```{r plotLocalAuthorities}
ggplot()+
  geom_sf(data=Background %>% filter(Name!="Ireland"), aes(geometry=geom)) +
  geom_sf(data=Areas %>% filter(RegionNation != "Northern Ireland"), 
          aes(geometry=geom), colour="Black", size=0.2)+
  geom_sf(data=Groups %>% filter(RegionNation!="Northern Ireland"), 
          aes(geometry=geom), fill=NA, colour="Black") +
  geom_sf_text(data=Group_labels, aes(geometry=geom, label=Group.labe,
                                      hjust=just), size=rel(2.4), colour="Black") +
  theme_void() +
  theme(plot.title=element_text(face="bold", size=rel(1.5)),
        legend.position="top") +
  labs(title="Local Authorities Hex-Cartogram",
       caption="Cartogram by House of Commons Library\nPlot by @tom_rushby")
```

## Middle-layer Super Output Areas

Load GeoPackage file ... this time Middle-layer Super Output Areas (MSOAs) in England and Wales only.

```{r}
msoa <- tempfile()
source <- ("https://github.com/houseofcommonslibrary/uk-hex-cartograms-noncontiguous/raw/main/geopackages/MSOA.gpkg")
msoa <- curl_download(url = source, destfile = msoa, quiet = FALSE, mode = "wb")
```

List layers ...

```{r}
st_layers(msoa)
```

Extract layers ...

```{r}
background_msoa <- st_read(msoa, layer="5 Background")
msoa_la_outlines <- st_read(msoa, layer="3 Local authority outlines (2019)")
msoa_data <- st_read(msoa, layer="4 MSOA hex")

msoa_groups <- st_read(msoa, layer="2 Groups")

msoa_group_labels <- st_read(msoa, layer="1 Group labels") %>% 
  mutate(just=if_else(LabelPosit=="Left", 0, 1))
```

As with the local authority example, `msoa_groups` and `msoa_la_outlines` geometries provide County- and Local Authority-level grouping/outlines (black in the plot below). `msoa_data` provides the MSOA outlines (in blue).

```{r}
ggplot()+
  geom_sf(data=background_msoa %>% filter(Name!="Ireland"), aes(geometry=geom))+
  geom_sf(data=msoa_data %>% filter(RegionNation!="Northern Ireland"), 
          aes(geometry=geom), fill=NA, colour="Blue") +
  geom_sf(data=msoa_la_outlines %>% filter(RegionNation!="Northern Ireland"), 
          aes(geometry=geom), fill=NA, colour="Black") +
  geom_sf(data=msoa_groups %>% filter(RegionNation!="Northern Ireland"), 
          aes(geometry=geom), fill=NA, colour="Black") +
  geom_sf_text(data=msoa_group_labels, aes(geometry=geom, label=Group.labe,
                                      hjust=just), size=rel(2.4), colour="Black") +
  theme_void() +
  theme(plot.title=element_text(face="bold", size=rel(1.5)),
        legend.position="top") +
  guides(fill = guide_colorbar(title.position = 'top', title.hjust = .5,
                               barwidth = unit(20, 'lines'), barheight = unit(.5, 'lines'))) +
  labs(title="MSOA Hex-Cartogram",
       caption="Cartogram by House of Commons Library\nPlot by @tom_rushby")
```


